---
title: "Random Forest in R"
author: "Nick J Lyon"
engine: knitr
format: 
  revealjs: 
    slide-number: c
    scrollable: false
    code-overflow: wrap
    code-line-numbers: false
    code-copy: hover
    theme: [night, slides.scss]
    reference-location: document
---

## Prepare {.smaller}

- First, you'll need to install and load a few {{< fa brands r-project >}} R packages
    - While not technically necessary, the `librarian` package makes library management _much_ simpler

\

```{r libs}
#| echo: true

# Install librarian (if you need to)
# install.packages("librarian")

# Install (if not already present) and load needed libraries
librarian::shelf(tidyverse, randomForest, permimp, vegan)
```

## Lichen Data {.smaller}

- The `vegan` package includes some lichen community composition data we can use for exploratory purposes

\

- We'll begin by loading that data (with some minor wrangling)

```{r data-prep}
#| echo: true

# Load vegan's lichen dataset & associated chemistry dataset
utils::data("varespec", package = 'vegan')
utils::data("varechem", package = 'vegan')

# Get one lichen species' cover information separate
lichen_sp <- dplyr::select(varespec, Callvulg)

# Attach the single species to the chemistry data
lichen_df <- cbind(lichen_sp, varechem)
```

## Data Structure {.smaller}

- This data object now has the following structure:
```{r data-str}
#| echo: true

# Check lichen data structure
str(lichen_df)
```

## [Random Forest]{.edu-pink} {.smaller}

- Run the random forest with the function and package of the same name

```{r random-forest}
#| echo: true

# Actually do the random forest
lich_rf <- randomForest::randomForest(Callvulg ~ ., data = lichen_df, 
                                      ntree = 1000, mtry = 2, 
                                      na.action = na.omit,
                                      keep.forest = T, keep.inbag = T)
```

\

- Quick argument explanation
    - 'Y ~ .' format of model means all other columns are (potential) predictors
    - `ntree` is the number of trees in the forest
    - `mtry` is the number of variables per node in the tree

## Variable Importance Plot {.smaller}

- We can now generate a [variable importance plot]{.edu-gold} based on that random forest

```{r var-imp-plot}
#| echo: true
#| fig-align: center

# Create variable importance plot
randomForest::varImpPlot(x = lich_rf, sort = T,
                         n.var = (ncol(lichen_df) - 1),
                         main = "Variable Importance")
```

## Conditional Permutation Importance (CPI) {.smaller}

- We can use that random forest to perform conditional permutation

```{r cpi-1}
#| echo: true
#| fig-align: center
#| fig-height: 4

# Implement conditional permutation
high_thresh <- permimp::permimp(object = lich_rf, conditional = T,
                                # Note the threshold is set to 0.95
                               threshold = 0.95, do_check = F, progressBar = F)

# Make CPI plot
plot(high_thresh, type = "box", horizontal = T)
```

## CPI - Thresholds {.smaller}

- As you might imagine, the threshold you pick can have a dramatic effect!

```{r cpi-2}
#| echo: true
#| fig-align: center
#| fig-height: 4

# Implement conditional permutation
low_thresh <- permimp::permimp(object = lich_rf, conditional = T,
                               # Note the lower threshold
                               threshold = 0.50, do_check = F, progressBar = F)

# Make CPI plot
plot(low_thresh, type = "box", horizontal = T)
```

## Exploratory Plotting {.smaller}

- Let's graph the response against the four 'most important' variables
    - This part is just for fun!

```{r explore}
# Load the cowplot library
librarian::shelf(cowplot)

# First graph
g1 <- ggplot(lichen_df, aes(y = Callvulg, x = pH)) +
  geom_smooth(formula = "y ~ x", method = "lm", color = "black", se = F) +
  geom_point(color = "black", fill = "#bfd3e6", pch = 21, size = 3) +
  labs(x = "pH", y = "Callvulg Cover (%)") +
  supportR::theme_lyon()
  
# Second graph
g2 <- ggplot(lichen_df, aes(y = Callvulg, x = Al)) +
  geom_smooth(formula = "y ~ x", method = "lm", color = "black", se = F) +
  geom_point(color = "black", fill = "#8c96c6", pch = 22, size = 3) +
  labs(x = "Aluminum", y = "Callvulg Cover (%)") +
  supportR::theme_lyon()

# Third
g3 <- ggplot(lichen_df, aes(y = Callvulg, x = P)) +
  geom_smooth(formula = "y ~ x", method = "lm", color = "black", se = F) +
  geom_point(color = "black", fill = "#88419d", pch = 23, size = 3) +
  labs(x = "Phosphorus", y = "Callvulg Cover (%)") +
  supportR::theme_lyon()

# Final
g4 <- ggplot(lichen_df, aes(y = Callvulg, x = Baresoil)) +
  geom_smooth(formula = "y ~ x", method = "lm", color = "black", se = F) +
  geom_point(color = "black", fill = "#4d004b", pch = 24, size = 3) +
  labs(x = "Bare Soil", y = "Callvulg Cover (%)") +
  supportR::theme_lyon()

# Assemble and present
cowplot::plot_grid(g1, g2, g3, g4, ncol = 2, nrow = 2, labels = "AUTO")
```

## [Thanks! Questions?]{.cream} {background-image="images_general/pic-question-mark-bfly.jpg"}
